Ages of Planets and Stars Observed by Kepler

### 2024-04-16 paper/ms.tex
minor updates

\subsection{Gyrochronology Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select the subset stars for which we expect these age to in fact be
valid, we then build a set of quality flags.  When and
how these flags should be applied depends on the question being asked.
If the goal is to construct a false-positive free sample, they might
all be applied.
If the goal is constructing a complete sample of young stars,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  
The flags in Tables~\ref{tab:stars} and~\ref{tab:planets} enable
selecting for such cases using bitflags.

{\it Temperature range (bit 0)}---We required stars to have effective
temperatures $T_{\rm eff}$ between 3800 and 6200\,K
\citep{Bouma_2023}.   Stars hotter than 6200\,K spin down very slowly,
if at all, limiting the utility of rotation-based ages.  Stars cooler
than 3800\,K do seem to spin down over gigayear timescales (e.g. CITE
E.Newton thesis, the Engle paper, and Chiti's thing), but the
currently available open cluster data have yet to clarify at what
times, if any, the intrinsic scatter in the population decreases.

{\it Surface gravity (bit 1)}---We flagged stars as possible subgiants if they had
$\log g < 4.2$.  We set this threshold by examining the
relevant flag from \citet{berger_2018_radii_evolnstates},
cross-matched against \citet{Berger_2020a_catalog}.
\citet{berger_2018_radii_evolnstates} performed this classification
using the PARSEC evolutionary models, stellar radii, and DR25 stellar
properties catalog effective temperatures \citep{Mathur_2017}.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In instances
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun/}; we
used 4$''$ rather than the 1$''$ match because 91 KOIs are omitted in
the 1$''$ match, 23 of which have either ``candidate'' or
``confirmed'' status.  Almost all of these 23 are M dwarfs with high
proper motions.  Performing the same crossmatch at 4 arcseconds, only
two KOIs (KOI-3993 and KOI-6531), both known false positives, fail to
yield Gaia DR3 crossmatches.}, which is a crossmatch of the NASA
Exoplanet Archive \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia
DR3.  For KIC stars with multiple potential Gaia matches within the
4$''$ search radius, we adopted the brightest star as the default
match.  in most such cases this was unambiguous because there is a
large brightness difference between the primary and any apparent
neighbors.  Cases with multiple stars within 4$''$ within $\Delta
G$$<$$0.5$\,mag are noted using bit flag 5.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.

@@

{\it RUWE (bit 7)}---Following recommendations in the Gaia DR3
documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
we flagged stars with a Gaia DR3 renormalized unit weight error (RUWE)
exceeding 1.4 as possible binaries.  Such astrometric outliers can be
either bonafide astrometric binaries, or more often are marginally
resolved point sources for which the single-source PSF model used in
the Gaia pipeline induces spurious astrometric scatter.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This poses a
problem for rotation-based age-dating because {\it i)} ``stars'' far
from the main-sequence include unresolved and unidentified photometric
binaries, and {\it ii}) rotation periods of single stars far from the
main-sequence have likely evolved due to effects other than magnetic
braking.  The added complications of reddening and metallicity make
imposing a simple cut in these diagrams challenging.  After exploring
various options, we found that the $+1$$\sigma$ relative isochronal
uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{\rm
B20,iso}/t_{\rm B20,iso}$, traced the relevant region of parameter
space quite well.
Another way of saying this is that gyrochronology is likely to be less
reliable for stars for which the upper limit on the relative
isochronal uncertainty is small.

%FIXME TODO LEFT OFF HERE...

df['b20t2_rel_E_Age'] = np.abs(df['b20t2_E_Age'])/df['b20t2_Age']

This is potentially an issue for rotation-based age dating,
particularly

using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
loci to flag over or under-luminous stars in each diagram, typically
more than $\approx$0.5 magnitudes above or below the main sequence.
%TODO refer to a figure!!!

{\it Final calibration sample}---{\bf todo: mention any extra flags
from Berger2020, or perhaps in the KOI catalog itself, or perhaps the
amplitude cuts you might want to apply to the rotation period catalog}

@@

The rotation periods for the stellar and planetary samples are shown
in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods yielded
\nuniqstarsantosrot.  Further requiring these rotating stars
to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
over which \texttt{gyro-interp} returns finite age posteriors,
yields \nuniqstarsantosrotteffcut\ stars.
Further requiring them to be apparently single, near the main
sequence, with $\log g$$>$4.2, and not flagged as
eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
potentially amenable for gyrochronology.
Finally, imposing RUWE$<$1.4 leaves \nuniqstarsantosrotgyroappl\ stars
for which gyrochronology is likely to be valid.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Age distribution of stars and planet-hosts in the Kepler field
    inferred from rotation periods}.
    A histogram of samples from the age posteriors for all Kepler
    stars with reported rotation periods and temperatures is shown in
    light gray;
    the age calculation was actually valid for stars that meet the criteria
    discussed in Section~\ref{subsec:flags} (dark blue).
		\label{fig:hist_tgyro}
	}
\end{figure*}

@@

Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
transiting planets at 3.0, 7.8, and 58\,days, has a peculiar
architecture.  \citet{2019A&A...624A..15S} conducted a transiting
timing analysis of the system, confirming Kepler-411d and arguing for
the presence of a fourth non-transiting planet, Kepler-411e, based on
the TTVs seen in Kepler-411d and assuming that the system is nearly
coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
the \citet{2007ApJ...669.1167B} relation and the measured rotation
period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
star's temperature ($\approx$4900\,K) puts it in a regime where the
\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
observations for quite some time
\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.

Kepler-63b is interesting in that it is {\it missing} from our
catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
a polar orbit based on both the Rossiter-McLaughlin effect and a
starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
qualitatively consistent with the estimates from
\citet{2013ApJ...775...54S}, though with larger uncertainties because
our age accounts for the intrinsic scatter in the open cluster
calibration sequences \citep{Bouma_2023}.  
While the usual catalogs (\citetalias{Santos_2019},
\citetalias{Santos_2021}, \citetlias{McQuillan_2014}, and
\citetalias{Mazeh_2015}) do not report this star's rotation period.
\citetalias{Santos_2021} flagged the system as ``potentially'' having
rotation-based modulation.
{\bf todo: how many comparable missed cases are there?}.
%TODO FIXME it's bc you need to do mcquillan 2013....
%
%... David+2021 used:
% Angus+2018... McQuillan+2013... Mazeh+2015... Walkowicz&Basri2013...

>>>>

\subsection{Gyrochronology Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select stars for which we expect these age to in fact be valid, we
then build a set of quality flags.  When and how these flags should be
applied depends on the question being asked.  If the goal is to
construct a false-positive free sample, they might all be applied.  If
the goal is constructing a complete sample of young stars, then
consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  The flags in
Tables~\ref{tab:stars} and~\ref{tab:planets} enable selecting for or
against such cases using bit flags.

{\it Temperature range (bit 0)}---We require stars to have effective
temperatures $T_{\rm eff}$ between 3800 and 6200\,K in order to report
gyrochrone ages \citep{Bouma_2023}.   Stars hotter than 6200\,K spin
down very slowly, if at all.  Stars cooler than 3800\,K do spin down
over gigayear timescales (e.g. {\bf CITE} E.Newton thesis, the Engle
paper, and Chiti's article), but the currently available open cluster
data have yet to clarify at what times, if any, the intrinsic scatter
in the population decreases.

{\it Surface gravity (bit 1)}---We flagged stars as possible subgiants
if they had $\log g < 4.2$.  We set this threshold by examining the
relevant flag from \citet{berger_2018_radii_evolnstates},
cross-matched against \citet{Berger_2020a_catalog}.
%\citet{berger_2018_radii_evolnstates} performed this classification
%using the PARSEC evolutionary models, stellar radii, and DR25 stellar
%properties catalog effective temperatures \citep{Mathur_2017}.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun}, last
accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
because 91 KOIs are omitted in the 1$''$ match, 23 of which have
either ``candidate'' or ``confirmed'' status.  Almost all of these 23
are M dwarfs with high proper motions.  Performing the same crossmatch
at 4 arcseconds, only two KOIs (KOI-3993 and KOI-6531), both known
false positives, fail to yield Gaia DR3 crossmatches.}, which is a
crossmatch of the NASA Exoplanet Archive
\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ search radius, we adopted the brightest star
as the default match.  In most such cases this was unambiguous because
there is a large brightness difference between the primary and any
apparent neighbors.  However cases with multiple stars within 4$''$
within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

{\it RUWE (bit 7)}---We flagged stars with a Gaia DR3 renormalized
unit weight error
(RUWE\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
exceeding 1.4 as possible binaries.  Such astrometric outliers can be
either bona fide astrometric binaries, or more often are marginally
resolved point sources for which the single-source PSF model used in
the Gaia pipeline induces spurious astrometric scatter.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This is a
problem for rotation-based age-dating because {\it i)} ``stars'' far
from the main-sequence include unresolved photometric binaries, and
{\it ii}) rotation periods of single stars far from the main-sequence
have likely evolved due to effects other than magnetic braking.  The
added complications of reddening and metallicity make imposing a
simple cut in any two dimensions challenging.  After exploring various
options, we found however that the $+1$$\sigma$ relative isochronal
uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{\rm
B20,iso}/t_{\rm B20,iso}$, reasonably separated the relevant region of
parameter space.  Another way of saying this is that gyrochronology is
probably less reliable for stars for which the upper limit on the
relative isochronal uncertainty is small.
We flag stars with $+\sigma_{\rm B20,iso}/t_{\rm B20,iso} < 0.4$ as
meeting this criterion.

{\it Candidate pulsators and close-in binaries (bit
10)}---\citeauthor{Santos_2021} included in a flag for candidate
``classical pulsators'' (e.g. RR~Lyrae, Cepheids) and candidate
close-in binaries.  Visual inspection hints that this flag also
includes many bona fide young stars, and so although we propagate it 
to our set of quality flags, we generally ignore it.

An important decision for our subsequent analysis is how to combine
the above flags in order to decide whether or not a star is suited for
gyrochronology.  By default, we assume that a star is suitable for
gyrochronology if none of bits zero through nine are raised.

@@

The rotation periods for the stellar and planetary samples are shown
in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods and a
reliable DR3 crossmatch yielded
\nuniqstarsantosrot\ stars.  Further requiring these rotating stars
to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
over which \texttt{gyro-interp} returns finite age posteriors,
yields \nuniqstarsantosrotteffcut\ stars.
Further requiring them to be apparently single, near the main
sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
potentially amenable for gyrochronology.
Finally, imposing RUWE$<$1.4 and requiring no possible blends ($\Delta
G$<2.5) within 4$''$ leaves \nuniqstarsantosrotgyroappl\ stars for
which gyrochronology is likely to be valid.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Age distribution of stars (left) and known planet hosts
    (right) in the Kepler field inferred from rotation periods}.
    A histogram of samples from the age posteriors for all Kepler
    stars with reported rotation periods and temperatures is shown in
    light gray;
    the age calculation was actually valid for stars that meet the criteria
    discussed in Section~\ref{subsec:flags} (dark blue).
		\label{fig:hist_tgyro}
	}
\end{figure*}

@@

Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
transiting planets at 3.0, 7.8, and 58\,days, has a rare
architecture.  \citet{2019A&A...624A..15S} conducted a transiting
timing analysis of the system, confirming Kepler-411d and arguing for
the presence of a fourth non-transiting planet, Kepler-411e, based on
the TTVs seen in Kepler-411d and assuming that the system is nearly
coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
the \citet{2007ApJ...669.1167B} relation and the measured rotation
period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
star's temperature ($\approx$4900\,K) puts it in a regime where the
\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
observations for quite some time
\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.

Kepler-63b is interesting in that it is {\it missing} from our
catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
a polar orbit based on both the Rossiter-McLaughlin effect and a
starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
qualitatively consistent with the estimates from
\citet{2013ApJ...775...54S}, though with larger uncertainties because
our age accounts for the intrinsic scatter in the open cluster
calibration sequences \citep{Bouma_2023}.  
While the usual catalogs (\citetalias{Santos_2019},
\citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
\citetalias{Mazeh_2015}) do not report this star's rotation period.
\citetalias{Santos_2021} flagged the system as ``potentially'' having
rotation-based modulation.
{\bf todo: how many comparable missed cases are there?}.
%TODO FIXME it's bc you need to do mcquillan 2013....
%
%... David+2021 used:
% Angus+2018... McQuillan+2013... Mazeh+2015... Walkowicz&Basri2013...
